Renewal and stability in populations structured by remaining years of life

### 2014-06-05 ThanoRepro/ThanoRepro.tex
incorporated Brouard refs.

All demographic forces vary over age in known and regular ways. Information on
such forces, along with the size and age structure of a population allows the
demographer to make predictions about the future size and structure of a population. In this
paper, I explore some formal demographic consequences of a particular
redefinition of age. Instead of counting age as the time passed between birth and the present (or
some other moment), consider age as the amount of time left from the present
until death. Individuals in this case move in the same direction along
imaginary life lines, but the reference point for age is now at the end of the
life line instead of at the beginning. Of course, such information cannot be
requested of survey respondents,
because one's future time of death is typically unknown. It would be
possible to organize lifecourse data by remaining years of life in a
very long running linked population register, but such registers are few and far
between, and the analysis would be very retrospective in nature.
Instead we approximate this perspective using a strategy similar to that used by
\citet{miller2001increasing} or \citet{lee2002approach}, which uses lifetables
and aggregated population counts to redistribute counts within age-classes over
future death times. This allows us to make statements about the remaining
lifelines within a current stock of population, under certain assumptions.

@@

Before hashing out the thanatological renewal model, I present the
underlying data transformation --- a little formula that
gives the probability of dying at some age $y$ years in the future, given survival to some earlier age, $a$. For age 0 this probability is simply the $d(a)$ column of the lifetable, standardized to sum to 1. For higher ages, we simply condition on survival:

@@

We refer to \eqref{eq:transform} as the thanatological age
transformation\footnote{I have on good authority that Ken Wachter
coined the term \textit{thanatological} age.}, and this complements the notion
of chronological age. It bears repeating that thanatological age is
probabilistic until after death, since it refers to the future according to
particular assumptions about mortality. As such, Figure~\ref{fig:USrecomp} is
projective in nature, a forward-looking glance at a population attrition given the age-structure and
lifetable of a particular moment, whereas Figure~\ref{fig:USdecomp} is reflective in nature, since a population's
chronological age-structure is mostly the fruit of past fertility, but also
migration and to a lesser degree attrition. I explore population structure by remaining years of life
for its own sake in a different paper.\footnote{Poster Session 6 (Friday
mid-day))} Note that any count classified by chronological age can be
reclassified in this way, as long as you have a lifetable that plausibly represents the population whose data you wish to reclassify. This means we can derive thanatological fertility rates, $F_y$, by applying \eqref{eq:transform}
to age-classified birth counts, $B_a$, to get $B_y$ and again to
exposure-to-risk, $E_a$ (here we take exposures from all ages), to get $E_y$ and then dividing:

>>>>

All demographic forces vary over age in known and regular ways. Information on
such forces, along with the size and age structure of a population allows the
demographer to make predictions about the future size and structure of a population. In this
paper, I explore some formal demographic consequences of a particular
redefinition of age. Instead of counting age as the time passed between birth and the present (or
some other moment), consider age as the amount of time left from the present
until death\footnote{This quantity is referred to as \textit{residual} life in
the reliability literaure.}.
Individuals in this case move in the same direction along imaginary life lines, but the reference point for age is now at the end of the
life line instead of at the beginning. Of course, such information cannot be
requested of survey respondents,
because one's future time of death is typically unknown. It would be
possible to organize lifecourse data by remaining years of life in a
very long running linked population register, but such registers are few and far
between, and the analysis would be very retrospective in nature.
Instead we approximate this perspective using a strategy similar to that used by
\citet{brouard1986structure}, \citet{miller2001increasing}, or
\citet{lee2002approach}, which uses lifetables and aggregated population counts to redistribute counts within age-classes over
future death times. This allows us to make statements about the remaining
lifelines within a current stock of population, under certain assumptions.

@@

Before hashing out the thanatological renewal model, I present the
underlying data transformation --- a little formula that
gives the probability of dying at some age $y$ years in the future, given
survival to some earlier age, $a$. For age 0 this probability is simply the $d(a)$ column of the
lifetable, standardized to sum to 1. For higher ages, we simply condition on survival:\footnote{In the reliability literature, this
our concept of $d(y|a)$ is denoted by \begin{equation}
X - t | X > t
\end{equation}
where $X$ is the total lifespan and $t$ is the current age attained.}

@@

We refer to \eqref{eq:transform} as the thanatological age
transformation\footnote{I have on good authority that Ken Wachter
coined the term \textit{thanatological} age.}, and this complements the notion
of chronological age. \citep{brouard1989mouvements}, and later
\citep{vaupel2009life}, showed that in stationary populations the chronological
and thanatological age structures are identical.
It bears repeating that thanatological age is probabilistic until after death, since it refers to the future according to
particular assumptions about mortality. As such, Figure~\ref{fig:USrecomp} is
projective in nature, a forward-looking glance at a population attrition given the age-structure and
lifetable of a particular moment, whereas Figure~\ref{fig:USdecomp} is reflective in nature, since a population's
chronological age-structure is mostly the fruit of past fertility, but also
migration and to a lesser degree attrition. I explore population structure by remaining years of life
for its own sake in a different paper.\footnote{Poster Session 6 (Friday
mid-day))} Note that any count classified by chronological age can be
reclassified in this way, as long as you have a lifetable that plausibly represents the population whose data you wish to reclassify. This means we can derive thanatological fertility rates, $F_y$, by applying \eqref{eq:transform}
to age-classified birth counts, $B_a$, to get $B_y$ and again to
exposure-to-risk, $E_a$ (here we take exposures from all ages), to get $E_y$ and then dividing:
